UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-08 ms/main.tex
type in figure reference
add text on caveats

\subsubsection*{Caveats}
The existence of a (small) number of adjustable parameter to UMAP, including PC cutoff and number of neighbors, opens the door to tweaking. Figure~\ref{fig:supp_megamontage_pc10_50} shows that disconnected population clusters can appear at different locations depending on parameter choices. In the absence of clear theoretical rationales, we generally recommend to use as many PCs as are available and computationally feasible, and to use the default number of neighbours (15). We sometimes found that lower number of PCs led to a simpler shape that facilitated discussion (e.g., figure \ref{fig:fig_ukbb}). We recommend reporting on a range of parameter values and following up on observations with statistical testing. Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.

UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{images/1KGP_UMAP_PCS100_PCE3450_NC2_NN15_MD05_201944185843.jpeg}
    \end{subfigure}
    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in figure~\ref{fig:subpops}.}
    \label{fig:supp_1kgp_3350}
\end{figure*}

>>>>

\subsubsection*{Caveats}
The existence of adjustable parameter, including PC cutoff, minimum distance, and number of neighbors, opens the door to tweaking. Figure~\ref{fig:supp_megamontage_pc10_50} shows that disconnected population clusters can appear at different locations depending on parameter choices. Using a minimal number of neighbours (e.g. 5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact and losing visual detail, as well as possibly becoming disjoint. We sometimes found that a lower number of PCs led to a simpler shape that facilitated discussion (e.g. figure \ref{fig:fig_ukbb}). In the absence of clear theoretical rationales, we suggest to use as many PCs as are available and computationally feasible, and beginning with a minimum distance of 0.1 to 0.5 and the default number of neighbours (15). We recommend reporting on a range of parameter values and following up on observations with statistical testing.

Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations. UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{images/1KGP_UMAP_PCS100_PCE3450_NC2_NN15_MD05_201944185843.jpeg}
    \end{subfigure}
    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in figure~\ref{fig:1000g_compare}.}
    \label{fig:supp_1kgp_3350}
\end{figure*}
